---
title: "STAT547O - Bootstrap notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=6, 
                      message=FALSE, warning=FALSE)
```

#### LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 


# DRAFT (Read at your own risk)

## Bootstrap

Here is a simple example illustrating the use of the non-parametric bootstrap (Efron 1979) to
estimate the (joint) distribution of the estimated eigenvalues of the covariance
matrix of a random vector. To simplify the presentation we consider here a bi-variate
case `p = 2`). 

We will generate a random sample of `n = 200` observations from a 
2-dimensional Gaussian random vector with mean 0. We also create a 
covariance matrix at random (by generating a random matrix of eigenvectors)
and constructing `\Sigma = U \Lambda U^T`, where `\Lambda = diag(12, 3)`. 

```{r block1}
p <- 2
# create a population covariance matrix for the example
set.seed(123456)
u <- qr.Q(qr(matrix(rnorm(p*p), p, p)))
la <- c(12, 3) # c(3, 1)
( sigma <- u %*% diag(la) %*% t(u) )
```

We now generate the random sample:
```{r block2}
n <- 200
set.seed(123456)
x <- MASS::mvrnorm(n=n, mu=rep(0,p), Sigma=sigma)
```
The estimated eigenvalues of the true covariance matrix `\Sigma` are
```{r block3}
( la.hat <- svd(cov(x))$d )
```
Can we construct 95% confidence intervals for the each of the true eigenvalues? Or
a 95% confidence region for both of them simultaneously?

What is the distribution of the random vector `(\hat{\lambda}_1, \hat{\lambda}_2)^T` ? 

We will use the non-parametric bootstrap (Efron 1979) to estimate this distribution.
Our distribution estimator is the "empirical" (really: Monte Carlo) distribution obtained
by re-computing the estimator on samples obtained from the original sample. In this
example we will use `B = 500` bootstrap samples.

For each sample we will compute the eigenvalues of the corresponding sample
covariance matrix, so we need space to save 500 vectors of dimension 2
(this is the matrix `la.hat.b` in the code chunk below). We
then draw a sample, compute the 
eigenvalues of the sample
covariance matrix and store them. 
```{r other}
B <- 500
la.hat.b <- array(0, dim=c(B, p))
set.seed(123456)
for(j in 1:B) {
  ii <- sample(n, repl=TRUE)
  la.hat.b[j, ] <- svd( cov(x[ii,]) )$d
}
```
These 500 samples are our distribution estimator. This is
how they look like. The red point is the 
center of the estimated distribution, and the light red
point is the original vector of estimated eigenvalues. 
```{r bootsam}
plot(la.hat.b, pch=19, col='gray')
la.mu <- colMeans(la.hat.b)
points(la.mu[1], la.mu[2], col='red', pch=19, cex=1.2)
points(la.hat[1], la.hat[2], col='tomato', pch=19, cex=1.2)

```

This distribution looks fairly elliptical (and Gaussian). 
We can draw a 95% confidence interval using this. The
light blue point is the true vector of eigenvalues.
```{r bootsam2}
aa <- var(la.hat.b) 
plot(la.hat.b, pch=19, col='gray')
xx <- ellipse::ellipse(aa, centre=la.mu)
lines(xx, cex=.7, col='darkgreen', lwd=3)
points(la.mu[1], la.mu[2], col='red', pch=19, cex=1.2)
points(la[1], la[2], col='steelblue', pch=19, cex=1.5)
```

If we did marginal confidence intervals, we would use quantiles of 
the estimated marginal distributions
```{r marg1}
qs <- as.numeric( quantile(la.hat.b[,1] - la.mu[1], c(.025, .975)) )
c(la.mu[1] - qs[2], la.mu[1] - qs[1])

qs <- as.numeric( quantile(la.hat.b[,2] - la.mu[2], c(.025, .975)) )
c(la.mu[2] - qs[2], la.mu[2] - qs[1])
```
Another way to look at it, using boxplots:
```{r marg2}
boxplot(la.hat.b, col='gray80')
points(1, la[1], pch=19, cex=1.75, col='steelblue')
points(2, la[2], pch=19, cex=1.75, col='steelblue')
```

**Spoiler alert**: we do know the asymptotic distribution of each
eigenvalue estimator ( `N(\lambda_j, 2*\lambda_j^2 / n)` ). 

**Sanity check**: 
```{r sanity}
diag( var(la.hat.b) )
2*la.hat^2/n
```

The asymptotic CI would have been
```{r asym1}
c(la.hat[1] - qnorm(.975)*sqrt(2*la.hat[1]^2/n), 
  la.hat[1] + qnorm(.975)*sqrt(2*la.hat[1]^2/n) )

c(la.hat[2] - qnorm(.975)*sqrt(2*la.hat[2]^2/n), 
  la.hat[2] + qnorm(.975)*sqrt(2*la.hat[2]^2/n) )
```



